---
title: ""
output: html_document
---

# Problem

# Data Selection
This data is slightly pre-cleaned and obtained through NASA APIs obtained through the Center for Near Earth Object Studies which is responsible for computing highly accurate orbital data for thousands of asteroids and comets that fly close to our planetary neighborhood. This cleaned version is available on Kaggle under public domain provided by Mr. Sameep Vani, and it was last updated in June 2022.

# Data Exploration

## Reading the Dataset
```{r}
data <- read.csv("neo_v2.csv")
```
## Load necessary libraries
```{r}
library(dplyr)
```

## Size of the Dataset
```{r}
nrow(data)
```
We can see we have 90,836 rows.

```{r}
length(unique(data$id))
```
That said, it's apparent that we only have 27,423 unique objects tracked in this dataset. Meaning certain objects appear more than once as they have orbited the Earth along the years.

## Examining Our Features
Let's now take a look to see what features are present for us in the dataset.
```{r}
columns <- colnames(data)
columns
```

So, we have 10 columns in our data. We have compiled a description of these columns in the table below:

| Column Name        | Data Type | Description                                                                        | Unit                |
|--------------------|-----------|------------------------------------------------------------------------------------|---------------------|
| id                 | int       | Unique identifier for each asteroid                                                | N/A                 |
| name               | str       | Name assigned to the object by NASA                                                | N/A                 |
| est_diameter_min   | float     | Minimum estimate of the diameter of the object                                     | kilometers          |
| est_diameter_max   | float     | Maximum estimate of the diameter of the object                                     | kilometers          |
| relative_velocity  | float     | Velocity of the object relative to Earth                                           | kilometers per hour |
| miss_distance      | float     | Distance missed                                                                    | kilometers          |
| orbiting_body      | str       | Planet that the object orbits                                                      | N/A                 |
| sentry_object      | bool      | Whether the object is included in the sentry-automated collision monitoring system | N/A                 |
| absolute_magnitude | float     | Describes the magnitude of the asteroid based on its intrinsic luminosity          | N/A                 |
| hazardous          | bool      | Whether or not the object is harmful                                               | N/A                 |

## Examining the Distribution of Values
We will now try to visualize the distribution of values in each column of our dataset to get a better understanding of the nature of the data.

```{r}
library(ggplot2)
```

### `est_diameter_min`
```{r}
# Create a density plot
ggplot(data, aes(x = est_diameter_min)) +
  geom_density() +
  labs(x = "Estimated Diameter (Min)", y = "Density", title = "Density Plot of Minimum Estimated Diameter")
```

### `est_diameter_max`
```{r}
ggplot(data, aes(x = est_diameter_max)) +
  geom_density() +
  labs(x = "Maximum Estimated Diameter", y = "Density", title = "Density Plot of Maximum Estimated Diameter")
```

### `relative_velocity`
```{r}
ggplot(data, aes(x = relative_velocity)) +
  geom_density() +
  labs(x = "Relative Velocity", y = "Density", title = "Density Plot of Relative Velocity")
```

### `miss_distance`
```{r}
ggplot(data, aes(x = miss_distance)) +
  geom_density() +
  labs(x = "Miss Distance", y = "Density", title = "Density Plot of Miss Distance")
```

### `orbiting_body`
```{r}
orbit_cat_count <- table(data$orbiting_body)
pie_chart <- pie(orbit_cat_count, labels = paste(names(orbit_cat_count), "\n", orbit_cat_count), main = "Distribution of Orbiting Bodies")

```

### `sentry_object`
```{r}
sentry_cat_count <- table(data$sentry_object)
pie_chart <- pie(sentry_cat_count, labels = paste(names(sentry_cat_count), "\n", sentry_cat_count), main = "Distribution of Sentry Objects")
```

### `absolute_magnitude`
```{r}
ggplot(data, aes(x = absolute_magnitude)) +
  geom_density() +
  labs(x = "Absolute Magnitude", y = "Density", title = "Density Plot of Absolute Magnitude")
```

### `hazardous`
```{r}
hazard_cat_counts <- table(data$hazardous)
pie_chart <- pie(hazard_cat_counts, labels = paste(names(hazard_cat_counts), "\n", hazard_cat_counts), main = "Distribution of Hazardous Objects")
```

## Under-sampling
Removing Duplicate IDs 
```{r}
# Remove duplicates based on ID
data1 <- data 

# Splitting the data into hazardous and non-hazardous datasets
hazardous_data <- filter(data1, hazardous == TRUE)
non_hazardous_data <- filter(data1, hazardous == FALSE)

# Calculate the total number of samples needed
total_samples <- nrow(data1)

# Since hazardous data is only 20% of the dataset, we use all of it
hazardous_samples <- nrow(hazardous_data)

# Calculate the number of non-hazardous samples needed to achieve the 55% ratio
# For this, first calculate the total desired size of the final dataset
final_dataset_size <- hazardous_samples / 0.45
non_hazardous_samples <- round(final_dataset_size * 0.55)

# Adjust in case the calculation exceeds the available non-hazardous data
non_hazardous_samples <- min(non_hazardous_samples, nrow(non_hazardous_data))

# Sample the non-hazardous data
sampled_non_hazardous <- sample_n(non_hazardous_data, non_hazardous_samples)

# Combine the samples (all hazardous data and sampled non-hazardous data)
final_dataset <- rbind(hazardous_data, sampled_non_hazardous)

# Shuffle the final dataset
set.seed(123)
final_dataset <- final_dataset[sample(nrow(final_dataset)),]

# Save the final dataset
write.csv(final_dataset, "balanced_dataset.csv", row.names = FALSE)

```


# Feature Engineering

## Encoding
```{r}
data$hazardous <- ifelse(data$hazardous == "True", 1, 0)
```

## Excluding Features

TO BE REPLACED BY PCA

As we saw in our data exploration step, some columns offer uninformative information. These columns are mainly `orbiting_body` and `sentry_object` as they have no variability at all.

Other columns we want to exclude will be the `name` and `id` as they are not even features, rather identifiers for the objects.

```{r}
feature_names <- c("absolute_magnitude", "miss_distance", "relative_velocity", "est_diameter_max", "est_diameter_min")
features <- data[feature_names]
target <- data$hazardous
```

# Data Preprocessing

## Train / Test Splits

# Model Development

## Decision Trees

### Fitting the Model

### Model Evaluation

### Pruning

## Support Vector Machines

## Principal Component Analysis

## Clustering

# Hyperparameter Tuning

# Interpreting the Results